Bucket-Sorted Independent Sets for Algebraic Multigrid

### 2007-09-24 bsis_paper/paper.tex
Overhauled Section 4, and wrote first content in the conclusions section (Section 5).
git-svn-id: file:///home/rblake/svn/pamg/trunk@29 c7f3fe13-8c30-0410-9127-9a9c42fe009c

For a general independent set-based coarse-grid selection algorithm, a
vertex $i$ is eligible to be added to $D$ if
\begin{equation}
\label{5:eqn:general-selection}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i^s,
\end{equation}
where $\mathcal{N}_i^s$ is the \emph{selection neighborhood} of
$i$. Although the distance-$d$ neighborhood is a sensible choice for
the selection neighborhood, this discussion is not limited to such
cases. It is assumed, however, the matrix formed by $\mathcal{N}_*^s$
(i.e., the selection sets for all vertices) is symmetric. This is
equivalent to stating if $j \in \mathcal{N}_i^s$, then $i \in
\mathcal{N}_j^s$ for all $i$ and $j$ in the vertex set. Furthermore,
this discussion assumes the weight of vertex $i$ is only decremented
following weight updates and is unaffected by the assignment of a
vertex $j \notin \mathcal{N}_i^s$ to the $C$-point set.

@@

When a vertex $i$ satisfies the generalized selection
condition~(\ref{5:eqn:general-selection}), no other valid $C$-point
assignments are able to affect $w_i$. This is formalized below.
\begin{lem}
\label{5:lemma:C-lock}
If (\ref{5:eqn:general-selection}) is satisfied for vertex $i$, then
$i$ must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become
a $C$-point, regardless of any other selections and updates made in
the graph.
\end{lem}
\begin{proof}
The satisfaction of (\ref{5:eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{5:eqn:general-selection}) and becomes
a $C$-point, which is impossible until after $i$ is assigned to the
$C$-point set.
\end{proof}

@@

Let $D_0$ be the set of all vertices satisfying
(\ref{5:eqn:general-selection}) in the first iteration of coarse-grid
selection. Vertices in $D_0$ become $C$-points regardless of the
algorithm used to build coarse grids. In most independent set-based
algorithms developed so far, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lem}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lem}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{5:eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}

@@

\begin{thm}
\label{5:theorem:invariance}
All independent set-based coarse-grid selection algorithms given the
same initial weights and using the same selection neighborhood,
selection criterion based on~(\ref{5:eqn:general-selection}), and
weight update heuristics as described above select identical coarse
grids.
\end{thm}
\begin{proof}
Let $c$ be the vertices in the extended selection neighborhood of $i$
($\mathcal{N}_i^{2s}$) satisfying the conditions for $D$ in some
arbitrary iteration. Suppose assigning vertices $c_1 \subset c$ to the
$C$-point set leads to $w_i > w_j$ for all $j \in
\mathcal{N}_i^s$. Also suppose assigning $c_2 \subset c$, $c_1 \ne
c_2$, to the $C$-point set leads to the existence of some $j \in
\mathcal{N}_i^s$ such that $w_j > w_k$ for all $k \in
\mathcal{N}_j^s$. In the first case, $i$ becomes a $C$-point before
$j$, and the opposite is true in the second case.

@@

\section{Bucket Sorted Independent Set Selection}
\label{sec:bsis}

\subsection{CLJP-c}
CLJP-c~\cite{alber-cljpc} colors the graph of $S$ before selecting
$C$-points, and the colors are used as one component of the vertex
weights. As a result, the structure of the coarse grids selected is
improved. More formally, the use of graph coloring provides the
following important result.
\begin{thm}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{thm}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{cor}
\label{cor:same-weight-indep-set}
Any set of vertices in the graph of $S$ sharing the same weight are an
independent set.
\end{cor}
\begin{cor}
\label{cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
vertex in that independent set has a uniquely maximal weight in its
neighborhood.
\end{cor}
Corollary~\ref{cor:same-weight-indep-set} states that independent sets
can be selected in the graph simply by selecting sets of vertices with
the same weight. Corollary~\ref{cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight,
updating weights, selecting the next set of vertices with maximum
weight, and so on. This approach is taken by the algorithm developed
in Section~\ref{sec:bsissub}. It is proven in
Section~\ref{5:sec:cg-invariance} that despite the difference in
approach, the resulting coarse grid is the same as that selected by
CLJP-c.

\subsection{BSIS}
\label{sec:bsissub}
In this section, new techniques for searching the graph of $S$ for new
$C$-points and subsequently updating the weights of remaining vertices
are developed. The new algorithm is named Bucket Sorted Independent
Sets (BSIS) to reflect the data structure used.

@@

\subsection{Motivation}
\label{sec:motivation}
CLJP and its descendents select an independent set $D$ in each
iteration. Recall, the condition for selecting $D$ is that all
vertices $i \in D$ must satisfy
\begin{equation}
\label{5:eqn:D-conditions}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i.
\end{equation}
CLJP and CLJP-c rely on a search routine to locate locally maximal
weights and on a weight update routine to modify weights of vertices
connected to new $C$-points.

The algorithms for searching and updating vertex weights in CLJP in
detail are examined in this section. In particular, the impact of
using a sparse matrix format on the coarsening procedure. The
pseudo-code below assumes the software uses a compressed sparse row
(CSR)~\cite{saadBook} matrix format or other similar format, which are
common matrix formats in numerical software. CSR provides low memory
costs for storing sparse matrices and provides efficient access to the
nonzeros in a row. Accessing the nonzeros in a column is an expensive
operation in this format and strongly influences the weight update
routine in CLJP-style algorithms because $S$ is, in general, not
symmetric.

The search step in CLJP is implemented as shown in
Algorithm~\ref{5:alg:search}.
\begin{lstlisting}[caption={Coarse-Grid Selection Graph Search},label=5:alg:search]
%\textsc{Search-Graph}$(S,\, C,\, F)$%
$D \leftarrow \emptyset$
for $i$ in $(C \cup F)$
  if $w_i > w_j,\,\forall j \in \mathcal{N}_i$
    $D \leftarrow D \cup \{i\}$
  end
end
return $D$
\end{lstlisting}
% \begin{algorithm}[h]
% \caption{Coarse-Grid Selection Graph Search}
% \label{5:alg:search}
% \textsc{Search-Graph}$(S,\, C,\, F)$ \{
% \begin{algorithmic}[1]
% \STATE $D \leftarrow \emptyset$
% \FORALL{$i \notin (C \cup F)$}
% \IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
% \STATE $D \leftarrow D \cup \{i\}$
% \ENDIF
% \ENDFOR
% \RETURN $D$
% \end{algorithmic}
% \}
% \end{algorithm}
In the first iteration, Line~3 is run $2|E|$ times. The total cost of
search in constructing the coarse grid depends on the number of
iterations needed. Even in the best case of $\Omega(E)$ time, the cost
is significant when the graph contains large numbers of edges, as
usually happens on the lower levels in the grid hierarchy
(see~\cite{alber-PCGS} for examples). In the next section, a new
technique is introduced for conducting the search in coarse-grid
selection algorithms independent of the number of edges in the graph.

Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{5:alg:setup}.
\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=5:alg:setup]
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
for $d \in D$
  for $i \in S_d$
    $w_i \leftarrow w_i - 1$    //see %Figure~\ref{3:fig:CLJP-update-heuristics} left%
    $S_d = S_d \setminus \{i\}$ //removing edge from graph
  end
end
for $i \notin (C \cup F)$
  for $k$ originally in $S_i$
    if $k \in D$
      mark $k$ NEW-C-POINT
    end
  end
  for $j \in S_i$
    if $j \notin D$
      for $k \in S_j$
        if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
          $w_j \leftarrow w_j - 1$        // see %Figure~\ref{3:fig:CLJP-update-heuristics} right%
          $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
        end
      end
    end
  end
  for $k$ originally in $S_i$
    if $k \in D$
      unmark $k$
      $S_i = S_i \setminus \{k\}$         // remove edge from $k$ to $i$, if present
    end
  end
end
\end{lstlisting}
%\begin{algorithm}[h]
%\caption{CLJP Weight Update for CSR Matrix}
%\label{5:alg:setup}
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
%\begin{algorithmic}[1]
%\FORALL{$d \in D$}
%\FORALL{$i \in S_d$}
%\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} left}
%\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
%\ENDFOR
%\ENDFOR
%\FORALL{$i \notin (C \cup F)$}
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE mark $k$ NEW-C-POINT
%\ENDIF
%\ENDFOR
%\FORALL{$j \in S_i$}
%\IF{$j \notin D$}
%\FORALL{$k \in S_j$}
%\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
%\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} right}
%\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
%\ENDIF
%\ENDFOR
%\ENDIF
%\ENDFOR
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE unmark $k$
%\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
%  if present}
%\ENDIF
%\ENDFOR
%\ENDFOR
%\end{algorithmic}
%\}
%\end{algorithm}
The level of complication in this update routine is due to the CSR
format and the need to find vertices strongly influenced by new
$C$-points. When a new $C$-point $k$ is selected, the first type of
weight update is trivial since determining the vertices in $S_k$ is
inexpensive. The second type of update is more expensive since the
vertices influenced by $k$ are difficult to determine in a CSR
format. The update requires a search of many vertices $i$ and all of
their strong influencing neighbors $j$. The routine then searches
strongly influencing $j$ to determine if any $k \in D$ strongly
influences both $i$ and $j$. The cost increases dramatically as the
density of $S$ increases. Large operator complexities have a
disproportionately large impact on coarse-grid selection run time. In
the next section, a modified update routine to compliment the new
search technique is introduced.

\subsection{Bucket Sorted Independent Sets Algorithm}
The BSIS algorithm creates a bucket data structure enabling a search
for new $C$-points without individually scanning each vertex and
associated edges. Figure~\ref{5:fig:bucket-algorithm} illustrates the
bucket data structure. The number of buckets in the data structure is
$\max_{i \in S}|S_i^T|$ times the number of colors in the graph. That
is, each possible weight in $S$ has its own bucket. The vertices are
distributed to the appropriate buckets during the setup of the
coarse-grid selection algorithm, where the bucket of a newly placed
vertex depends on the number of vertices it strongly influences and
its color. For example, Vertex~14 strongly influences six vertices and
is black. Therefore, it is placed into the black bucket in the sixth
group of buckets. More notably, the vertices in a bucket form an
independent set (e.g., vertices 4, 16, and 21).

In each iteration, the non-empty bucket with largest weight forms $D$
(see Corollary~\ref{cor:max-weight-indep-set}). These vertices are
assigned to the $C$-point set and removed from the data
structure. Vertex weight updates lead to corresponding updates to the
data structure, and new $F$-points are removed from the data
structure. These operations continue until all buckets are empty, at
which point the coarse-grid selection is
complete. Algorithms~\ref{5:alg:BSIS-setup},
\ref{5:alg:BSIS-indep-set-select}, and~\ref{5:alg:BSIS-weight-update}
outline the operations discussed above.
\begin{lstlisting}[caption={BSIS Data Structure Setup},label=5:alg:BSIS-setup]
%\textsc{BSIS-Setup}$(S)$%
for $i \in V$
  $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
  $bucket[bucketID].\textsc{Insert}(i)$
end
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{BSIS Data Structure Setup}
% \label{5:alg:BSIS-setup}
% \textsc{BSIS-Setup}$(S)$ \{
% \begin{algorithmic}[1]
% \FORALL{$i \in V$}
% \STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
% \STATE $bucket[bucketID].\textsc{Insert}(i)$
% \ENDFOR
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={Independent Set Selection},label=5:alg:BSIS-indep-set-select]
%\textsc{BSIS-Independent-Set-Selection}$(S)$%
return non-empty $bucket$ with largest $bucketID$
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{Independent Set Selection}
% \label{5:alg:BSIS-indep-set-select}
% \textsc{BSIS-Independent-Set-Selection}$(S)$ \{
% \begin{algorithmic}[1]
% \RETURN non-empty $bucket$ with largest $bucketID$
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={BSIS Weight Update},label=5:alg:BSIS-weight-update]
%\textsc{BSIS-Weight-Update}$(S)$%
$bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
$bucket[bucketID].\textsc{Remove}(i)$
$bucket[bucketID - numColors].\textsc{Insert}(i)$
\end{lstlisting}
%\begin{algorithm}[!h]
%\caption{BSIS Weight Update}
%\label{5:alg:BSIS-weight-update}
%\textsc{BSIS-Weight-Update}$(S)$ \{
%\begin{algorithmic}[1]
%\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
%\STATE $bucket[bucketID].\textsc{Remove}(i)$
%\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
%\end{algorithmic}
%\}
%\end{algorithm}
\begin{figure}
  \begin{center}
    %\includegraphics[width=0.8\textwidth]{images/bucket-algorithm}
    \includegraphics[width=0.98\textwidth]{images/BSIS/initial}
    \caption{The BSIS data structure.}
    \label{5:fig:bucket-algorithm}
  \end{center}
\end{figure}

Figure~\ref{5:fig:bucket-algorithm-cp1} illustrates the graph and data
structure following the first iteration of the algorithm. Vertex~10
has become a $C$-point and its neighbors weights have been
updated. Vertices assigned to $F$ or $C$ are removed from the data
structure and other affected vertices are moved to new locations in
the data structure. Such vertices are highlighted in red.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/BSIS/cp1}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10). The weights of neighbors of the new
    $C$-point are updated. Some neighbors become $F$-points and are
    removed from the data structure. Vertices removed from the data
    structure are highlighted with a red ring in the graph, while
    other neighbors are moved to new locations in the data structure
    and are highlighted (to help in reading the figure) in the data
    structure with a red box.}
    \label{5:fig:bucket-algorithm-cp1}
  \end{center}
\end{figure}

>>>>

For a general independent set-based coarse-grid selection algorithm, a
vertex $i$ is eligible to be added to $D$ if
\begin{equation}
\label{eqn:general-selection}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i^s,
\end{equation}
where $\mathcal{N}_i^s$ is the \emph{selection neighborhood} of
$i$. Although the distance-$d$ neighborhood is a sensible choice for
the selection neighborhood, this discussion is not limited to such
cases. It is assumed, however, the matrix formed by $\mathcal{N}_*^s$
(i.e., the selection sets for all vertices) is symmetric. This is
equivalent to stating if $j \in \mathcal{N}_i^s$, then $i \in
\mathcal{N}_j^s$ for all $i$ and $j$ in the vertex set. Furthermore,
this discussion assumes the weight of vertex $i$ is only decremented
following weight updates and is unaffected by the assignment of a
vertex $j \notin \mathcal{N}_i^s$ to the $C$-point set.

@@

When a vertex $i$ satisfies the generalized selection
condition~(\ref{eqn:general-selection}), no other valid $C$-point
assignments are able to affect $w_i$. This is formalized below.
\begin{lem}
\label{5:lemma:C-lock}
If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then
$i$ must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become
a $C$-point, regardless of any other selections and updates made in
the graph.
\end{lem}
\begin{proof}
The satisfaction of (\ref{eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{eqn:general-selection}) and becomes
a $C$-point, which is impossible until after $i$ is assigned to the
$C$-point set.
\end{proof}

@@

Let $D_0$ be the set of all vertices satisfying
(\ref{eqn:general-selection}) in the first iteration of coarse-grid
selection. Vertices in $D_0$ become $C$-points regardless of the
algorithm used to build coarse grids. In most independent set-based
algorithms developed so far, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lem}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lem}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}

@@

\begin{thm}
\label{5:theorem:invariance}
All independent set-based coarse-grid selection algorithms given the
same initial weights and using the same selection neighborhood,
selection criterion based on~(\ref{eqn:general-selection}), and weight
update heuristics as described above select identical coarse grids.
\end{thm}
\begin{proof}
Let $c$ be the vertices in the extended selection neighborhood of $i$
($\mathcal{N}_i^{2s}$) satisfying the conditions for $D$ in some
arbitrary iteration. Suppose assigning vertices $c_1 \subset c$ to the
$C$-point set leads to $w_i > w_j$ for all $j \in
\mathcal{N}_i^s$. Also suppose assigning $c_2 \subset c$, $c_1 \ne
c_2$, to the $C$-point set leads to the existence of some $j \in
\mathcal{N}_i^s$ such that $w_j > w_k$ for all $k \in
\mathcal{N}_j^s$. In the first case, $i$ becomes a $C$-point before
$j$, and the opposite is true in the second case.

@@

\section{Bucket-Sorted Independent Set Selection}
\label{sec:bsis}
In this section, we introduce Bucket-Sorted Independent Sets
(BSIS). BSIS uses the same policies as CLJP-c, so from the theory in
Section~\ref{sec:invariance}, BSIS and CLJP-c produce the same coarse
grids. BSIS is designed to be more efficient than CLJP-c.

\subsection{CLJP-c}
CLJP-c~\cite{alber-cljpc} colors the graph of $S$ before selecting
$C$-points, and the colors are used as one component of the vertex
weights. As a result, the structure of the coarse grids selected is
improved. More formally, the use of graph coloring provides the
following important result.
\begin{thm}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{thm}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{cor}
\label{cor:same-weight-indep-set}
Any set of vertices in the graph of $S$ sharing the same weight are an
independent set.
\end{cor}
\begin{cor}
\label{cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~(\ref{eqn:general-selection}). That is,
each vertex in that independent set has a uniquely maximal weight in
its neighborhood.
\end{cor}
Corollary~\ref{cor:same-weight-indep-set} states that independent sets
can be selected in the graph simply by selecting sets of vertices with
the same weight. Corollary~\ref{cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight in the
graph, updating weights, selecting the next set of vertices with
maximum weight, and so on. This approach is taken by the BSIS
algorithm.

CLJP and CLJP-c rely on a search routine to locate all vertices
satisfying~(\ref{eqn:general-selection}) (i.e., vertices with locally
maximal weights) and on a weight update routine to modify weights of
vertices in the neighborhoods of new $C$-points.

The algorithms for searching and updating vertex weights in CLJP in
detail are examined in this section. The pseudo-code below assumes the
software uses a compressed sparse row (CSR)~\cite{saadBook} matrix
format or other similar format, which are common matrix formats in
numerical software. CSR provides low memory costs for storing sparse
matrices and provides efficient access to the nonzeros in a
row. Accessing the nonzeros in a column is an expensive operation in
this format and strongly influences the weight update routine in
CLJP-style algorithms because $S$ is, in general, not symmetric.

The search step in CLJP is implemented as shown in
Algorithm~\ref{5:alg:search}.
\begin{lstlisting}[caption={Coarse-Grid Selection Graph Search},label=5:alg:search]
%\textsc{Search-Graph}$(S,\, C,\, F)$%
$D \leftarrow \emptyset$
for $i \notin (C \cup F)$
  if $w_i > w_j\ \textrm{for all } j \in \mathcal{N}_i$
    $D \leftarrow D \cup \{i\}$
  end
end
return $D$
\end{lstlisting}
% \begin{algorithm}[h]
% \caption{Coarse-Grid Selection Graph Search}
% \label{5:alg:search}
% \textsc{Search-Graph}$(S,\, C,\, F)$ \{
% \begin{algorithmic}[1]
% \STATE $D \leftarrow \emptyset$
% \FORALL{$i \notin (C \cup F)$}
% \IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
% \STATE $D \leftarrow D \cup \{i\}$
% \ENDIF
% \ENDFOR
% \RETURN $D$
% \end{algorithmic}
% \}
% \end{algorithm}
In the first iteration, Line~3 is run $2|E|$ times. The total cost of
search in constructing the coarse grid depends on the number of
iterations needed. Even in the best case of $\Omega(E)$ time, the cost
is significant when the graph contains large numbers of edges, as
usually happens on the lower levels in the grid hierarchy
(see~\cite{alber-PCGS} for examples). In the next section, a new
technique is introduced for conducting the search in coarse-grid
selection algorithms independent of the number of edges in the graph.

Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{alg:setup}.
\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=alg:setup]
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
for $d \in D$
  for $i \in S_d$
    $w_i \leftarrow w_i - 1$
    $S_d = S_d \setminus \{i\}$ //removing edge from graph
  end
end
for $i \notin (C \cup F)$
  for $k$ originally in $S_i$
    if $k \in D$
      mark $k$ NEW-C-POINT
    end
  end
  for $j \in S_i$
    if $j \notin D$
      for $k \in S_j$
        if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
          $w_j \leftarrow w_j - 1$
          $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
        end
      end
    end
  end
  for $k$ originally in $S_i$
    if $k \in D$
      unmark $k$
      $S_i = S_i \setminus \{k\}$         // remove edge from $k$ to $i$, if present
    end
  end
end
\end{lstlisting}
%\begin{algorithm}[h]
%\caption{CLJP Weight Update for CSR Matrix}
%\label{alg:setup}
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
%\begin{algorithmic}[1]
%\FORALL{$d \in D$}
%\FORALL{$i \in S_d$}
%\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} left}
%\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
%\ENDFOR
%\ENDFOR
%\FORALL{$i \notin (C \cup F)$}
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE mark $k$ NEW-C-POINT
%\ENDIF
%\ENDFOR
%\FORALL{$j \in S_i$}
%\IF{$j \notin D$}
%\FORALL{$k \in S_j$}
%\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
%\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} right}
%\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
%\ENDIF
%\ENDFOR
%\ENDIF
%\ENDFOR
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE unmark $k$
%\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
%  if present}
%\ENDIF
%\ENDFOR
%\ENDFOR
%\end{algorithmic}
%\}
%\end{algorithm}
The level of complication in this update routine is due to the CSR
format and the need to find vertices strongly influenced by new
$C$-points. When a new $C$-point $k$ is selected, the first type of
weight update (Lines 2--7) is trivial since determining the vertices
in $S_k$ is inexpensive. The second type of update (Lines 8--30) is
more expensive since the vertices influenced by $k$ are difficult to
determine in a CSR format. The update requires a search of many
vertices $i$ and all of their strong influencing neighbors $j$. The
routine then searches strongly influencing $j$ to determine if any $k
\in D$ strongly influences both $i$ and $j$. The cost increases
dramatically as the density of $S$ increases. Large operator
complexities have a disproportionately large impact on coarse-grid
selection run time. In the next section, a modified update routine to
compliment the new search technique is introduced.

\subsection{Bucket Sorted Independent Sets Algorithm}
\label{sec:bsis-alg}
In this section, new techniques for searching the graph of $S$ for new
$C$-points and subsequently updating the weights of remaining vertices
are developed. The new algorithm is named Bucket Sorted Independent
Sets (BSIS) to reflect the data structure used.

@@

Figure~\ref{5:fig:bucket-algorithm} illustrates the bucket data
structure. The number of buckets in the data structure is $\max_{i \in
S}|S_i^T|$ times the number of colors in the graph. That is, each
possible weight in $S$ has its own bucket. The vertices are
distributed to the appropriate buckets during the setup of the
coarse-grid selection algorithm, where the bucket of a newly placed
vertex depends on the number of vertices it strongly influences and
its color. For example, Vertex~14 strongly influences six vertices and
is black. Therefore, it is placed into the black bucket in the sixth
group of buckets. Notably, the vertices in a bucket form an
independent set (e.g., Vertices 14, 16, and 21).

In each iteration, the non-empty bucket with largest weight forms $D$
(see Corollary~\ref{cor:max-weight-indep-set}). These vertices are
assigned to the $C$-point set and removed from the data
structure. Vertex weight updates lead to corresponding updates to the
data structure, and new $F$-points are removed from the data
structure. These operations continue until all buckets are empty, at
which point the coarse-grid selection is
complete. Algorithms~\ref{5:alg:BSIS-setup},
\ref{5:alg:BSIS-indep-set-select}, and~\ref{5:alg:BSIS-weight-update}
outline the operations discussed above.
\begin{lstlisting}[caption={BSIS Data Structure Setup},label=5:alg:BSIS-setup]
%\textsc{BSIS-Setup}$(S)$%
for $i \in V$
  $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
  $bucket[bucketID].\textsc{Insert}(i)$
end
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{BSIS Data Structure Setup}
% \label{5:alg:BSIS-setup}
% \textsc{BSIS-Setup}$(S)$ \{
% \begin{algorithmic}[1]
% \FORALL{$i \in V$}
% \STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
% \STATE $bucket[bucketID].\textsc{Insert}(i)$
% \ENDFOR
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={Independent Set Selection},label=5:alg:BSIS-indep-set-select]
%\textsc{BSIS-Independent-Set-Selection}$(S)$%
return non-empty $bucket$ with largest $bucketID$
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{Independent Set Selection}
% \label{5:alg:BSIS-indep-set-select}
% \textsc{BSIS-Independent-Set-Selection}$(S)$ \{
% \begin{algorithmic}[1]
% \RETURN non-empty $bucket$ with largest $bucketID$
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={BSIS Weight Update},label=5:alg:BSIS-weight-update]
%\textsc{BSIS-Weight-Update}$(S,\, i)$%
$bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
$bucket[bucketID].\textsc{Remove}(i)$
$bucket[bucketID - numColors].\textsc{Insert}(i)$
\end{lstlisting}
%\begin{algorithm}[!h]
%\caption{BSIS Weight Update}
%\label{5:alg:BSIS-weight-update}
%\textsc{BSIS-Weight-Update}$(S)$ \{
%\begin{algorithmic}[1]
%\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
%\STATE $bucket[bucketID].\textsc{Remove}(i)$
%\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
%\end{algorithmic}
%\}
%\end{algorithm}
\begin{figure}
  \begin{center}
    %\includegraphics[width=0.8\textwidth]{images/bucket-algorithm}
    \includegraphics[width=0.98\textwidth]{images/BSIS/initial}
    \caption{The BSIS data structure.}
    \label{5:fig:bucket-algorithm}
  \end{center}
\end{figure}

Figure~\ref{5:fig:bucket-algorithm-cp1} illustrates the graph and data
structure following the first iteration of the algorithm. Vertex~10
has become a $C$-point and its neighbors' weights have been
updated. Vertices assigned to $F$ or $C$ are removed from the data
structure and other affected vertices are moved to new locations in
the data structure. Such vertices are highlighted in red.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/BSIS/cp1}
    \caption{The BSIS data structure following selection of the first
    $C$-point (Vertex 10). The weights of neighbors of the new
    $C$-point are updated. Some neighbors become $F$-points and are
    removed from the data structure. Vertices removed from the data
    structure are highlighted with a red ring in the graph, while
    other neighbors are moved to new locations in the data structure
    and are highlighted (to aid in reading the figure) in the data
    structure with a red box.}
    \label{5:fig:bucket-algorithm-cp1}
  \end{center}
\end{figure}

The weight update routine described in Algorithm~\ref{alg:setup} is
very expensive in this context because some iterations of BSIS select
few $C$-points. For a graph with a large number of colors, BSIS may
execute dozens or hundreds of low-cost iterations to select a coarse
grid. Recall the weight update routine described loops through all
unassigned vertices each time it is called, so when it is run by BSIS,
work is done on many unaffected vertices.

The largest factor in the cost of the weight update results from
searching for the second type of weight update in
Algorithm~\ref{alg:setup}, which is done by looping through all
unassigned vertices since a new $C$-point cannot easily determine
which vertices it strongly influences in a CSR matrix. It is less
expensive in this situation to construct the transpose of $S$ than to
search the entire graph in each iteration. In $S^T$, a $C$-point
quickly determines which vertices it influences and ``paints''
them. The algorithm then loops through all painted vertices and
determines if any are neighbors. This is a simple solution that has a
dramatic effect on the performance of BSIS, although the update cost
remains approximately equivalent to the cost in
CLJP-c.
